---
title: "Divvy_exersice"
author: "Suhendra"
date: '2022-03-28'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Divvy_Exercise_Full_Year_Analysis.
Note : This analysis is based on the Divvy case study "'Sophisticated, Clear, and Polished’: Divvy and Data Visualization" written by Kevin Hartman.

In this case study, I will perform data analysis for a fictional bike-share company in order to help them attract more riders.


## Cyclistic casual riders vs members.
Upload data, Cleaning, Merging, and Visualization the Data
link to data set.

Download the data set click here [link](https://divvy-tripdata.s3.amazonaws.com/index.html), I followed the script supplied in the Case Study 1 pdf [link](https://drive.google.com/drive/folders/1ZAhZ4YtseisIcSWqjNZOA9TQCa7JJI63?usp=sharing).


##  Setting up my environment
Note : Setting up my R environment by loading the 'tidyverse', 'ggplot2', 'lubridate' packages.

```{r}
library("tidyverse")
library(ggplot2)
library(lubridate)
```

## Upload Divvy dataset(csv file).

```{r}
q2_2019 <- read_csv("C:/data analyst/couse 8/DATA/Divvy_Trips_2019_Q2.csv")
q3_2019 <- read_csv("C:/data analyst/couse 8/DATA/Divvy_Trips_2019_Q3.csv")
q4_2019 <- read_csv("C:/data analyst/couse 8/DATA/Divvy_Trips_2019_Q4.csv")
q1_2020 <- read_csv("C:/data analyst/couse 8/DATA/Divvy_Trips_2020_Q1.csv")
```


## Compare data and combine into single file.

### Consistent column names
* Column names were not consistent between each quarter.
* Dataset base is q1_2020

```{r}
(q4_2019 <- rename(q4_2019,
                   ride_id = trip_id,
                   rideable_type =bikeid,
                   started_at = start_time,
                   ended_at = end_time,
                   start_station_name = from_station_name,
                   end_station_name = to_station_name,
                   start_station_id = to_station_id,
                   end_station_id = to_station_id,
                   member_casual = usertype))

(q3_2019 <- rename(q3_2019,
                   ride_id = trip_id,
                   rideable_type =bikeid,
                   started_at = start_time,
                   ended_at = end_time,
                   start_station_name = from_station_name,
                   end_station_name = to_station_name,
                   start_station_id = to_station_id,
                   end_station_id = to_station_id,
                   member_casual = usertype))

(q2_2019 <- rename(q2_2019,
                   ride_id = "01 - Rental Details Rental ID",
                   rideable_type = "01 - Rental Details Bike ID",
                   started_at = "01 - Rental Details Local Start Time",
                   ended_at = "01 - Rental Details Local End Time",
                   start_station_name = "03 - Rental Start Station Name",
                   end_station_name = "02 - Rental End Station Name",                  
                   start_station_id = "03 - Rental Start Station ID",
                   end_station_id = "02 - Rental End Station ID",                     
                   member_casual= "User Type"))
```

### Modified data types to match each other.
* Some columns referencing the same variables had different data types.

```{r}
q4_2019 <- mutate(q4_2019, ride_id = as.character(ride_id),
                    rideable_type = as.character(rideable_type))
  
q3_2019 <- mutate(q3_2019, ride_id = as.character(ride_id),
                    rideable_type = as.character(rideable_type))
  
q2_2019 <- mutate(q2_2019, ride_id = as.character(ride_id),
                    rideable_type = as.character(rideable_type))
  
q1_2020 <- mutate(q4_2019, ride_id = as.character(ride_id),
                    rideable_type = as.character(rideable_type))
  
```

### Combine the four databases into one big data frame.
* combine Q2_2019,Q3_2019,Q4_2018,and Q1_2020 into one file.

```{r}
all_trips <- bind_rows(q2_2019, q3_2019, q4_2019,q1_2020)
```

### Removing Unnecessary Columns
* Some columns are not needed and should be removed from the data frame

```{r}
all_trips <- all_trips %>%  
  select(-c(birthyear, gender, "01 - Rental Details Duration In Seconds Uncapped", "05 - Member Details Member Birthday Year", "Member Gender", "tripduration"))
```


## CLEAN UP AND ADD DATA TO PREPARE FOR ANALYSIS

### There are a few problems we will need to fix:

#### (1) In the "member_casual" column, there are two names for members ("member" and "Subscriber") and two names for casual riders ("Customer" and "casual"). We will need to consolidate that from four to two labels.

* In the "member_casual" column, replace "Subscriber" with "member" and "Customer" with "casual"

```{r}
all_trips <-  all_trips %>% 
  mutate(member_casual = recode(member_casual
                                ,"Subscriber" = "member"
                                ,"Customer" = "casual"))
```

* Display table were reassigned :

```{r}
table(all_trips$member_casual)
```

#### (2) The data can only be aggregated at the ride-level, which is too granular. We will want to add some additional columns of data -- such as day, month, year -- that provide additional opportunities to aggregate the data.

*  Spitting columns list into date, month, day, and year of each ride.

```{r}
all_trips$date <- as.Date(all_trips$started_at) #The default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```


#### (3) We will want to add a calculated field for length of ride since the 2020Q1 data did not have the "tripduration" column. We will add "ride_length" to the entire dataframe for consistency.

* Adding new column called "ride_length" to all_trips.

```{r}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)

```

* Convert ride length from factor to numeric so we can run calculations on the data.

```{r}
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(
  all_trips$ride_length))
is.numeric(all_trips$ride_length)
```

#### (4) There are some rides where tripduration shows up as negative, including several hundred rides where Divvy took bikes out of circulation for Quality Control reasons. We will want to delete these rides.

* Cleaning Bad Data.
Removed the dataframe ride_length was negative and create a new version of the dataframe (v2).

```{r}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]

```

## CONDUCT DESCRIPTIVE ANALYSIS.

### Sumarizing Aggregate Data & compare members and casual users.
Descriptive analysis on ride_length in mean, median, mid, max Casual vs. Member.

* straight average (total ride length / rides).

```{r}
aggregate(all_trips_v2$ride_length~all_trips_v2$member_casual, FUN = mean)

```

* midpoint number in the ascending array of ride lengths.

```{r}
aggregate(all_trips_v2$ride_length~all_trips_v2$member_casual, FUN = median)

```

* longest ride.

```{r}
aggregate(all_trips_v2$ride_length~all_trips_v2$member_casual, FUN = max)

```

* #shortest ride

```{r}
aggregate(all_trips_v2$ride_length~all_trips_v2$member_casual, FUN = min)

```


### Analyze ridership data by type and weekday

* Order the days of the week:

```{r}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week,levels=
  c("Sunday","Monday","Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

* Grouping By Weekday and User Type

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% #creates weekday field using wday()
  group_by(member_casual, weekday) %>% #groups by usertype and weekday
    summarise(number_of_rides = n(), #calculates the number of rides and average duration 
              average_duration = mean(ride_length))%>%  # calculates the average duration
  arrange(member_casual, weekday)	# sorts
```
 
### Visualize the number of rides by rider type.

* Visualization for number Of Riders by rider type (Weekdays).

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at,label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n(),
average_duration = mean(ride_length)) %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill= member_casual))+
           geom_col(position = "dodge")
```

 
* Visualization for Average duration(Weekdays).

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at,label = TRUE)) %>% 
  group_by(member_casual,weekday) %>% 
  summarise(number_of_rides = n(),
  average_duration = mean(ride_length)) %>% 
  ggplot(aes(x = weekday, y=average_duration, fill= member_casual))+
  geom_col(position = "dodge")
```

 
